---
title: "factors"
format: html
---

## factor documentation

time units:

-   pp   = prior period
-   cp   = current period
-   diff = difference between pp and cp

aggregate dimensions
-   sum  = sum the the value_var  
-   mean = take the simple row level average of the the value_var  
-   row  = keep the row level granularity without aggregating
-   wtdavg  = take the wtd. average of the value_var



```{r}

library(tidyverse)
devtools::load_all()

db <- fpaR::create_contonso_duckdb()

sales_db <- db$sales

sales_tbl <- 
  fpaR::sales |>
  mutate(
    month=month(order_date)
  ) |>
  group_by(
    month
    ,product_key
  ) |>
  summarize(
    quantity=sum(quantity)
    ,net_price=sum(net_price)
    ,.groups="drop"
  ) |>
  group_by(month) |>
  arrange(product_key,.by_group = TRUE) |>
  ungroup() |>
  mutate(
    price_realization=net_price/quantity
    ,mix=net_price/sum(net_price)
    ,revenue=net_price*quantity
  ) |>
  filter(
    month %in% c(1,2)
  ) |> 
    select(
    month,product_key,quantity,net_price,revenue
  ) 


sales_tbl |> 

  pivot_wider(
    names_from = month
    ,values_from=quantity:revenue
  )
  
  
sales_slim_tbl <- sales_tbl |> 
  group_by(month) |> 
  mutate(
    total_revenue=sum(revenue)
    ,total_quantity=sum(quantity)
  ) |> 
  group_by(product_key) |> 
  mutate(
    lag_quantity=lag(quantity,1,NA)
    ,lag_revenue=lag(revenue,1,NA)
    ,lag_net_price=lag(net_price,1,NA)
    ,lag_total_revenue=lag(total_revenue,1,NA)
     ,lag_total_quantity=lag(total_quantity,1,NA)
    ,delta_quantity=quantity-lag_quantity
    ,delta_net_price=net_price-lag_net_price
    ,delta_revenue=revenue-lag_revenue
    ,mix=revenue/total_revenue
    ,lag_max=lag_revenue/lag_total_revenue
    ,quantity_prop=quantity/total_quantity
    ,lag_quantity_prop=lag_quantity/lag_total_quantity
  ) |> 
  arrange(product_key) |> 
  drop_na()


## factor analysis
sales_slim_tbl |> 
  group_by(month) |> 
  mutate(
    volume_effect=((quantity*lag_quantity_prop)-lag_quantity)*net_price
  ) |> 
  ungroup() |> 
  summarise(
    volume_effect=sum(volume_effect)
  )
  relocate(volume_effect)
  
```


vol=(cp(quantity,row)*pp(quantity_prop,row)-pp(quantity-row))*cp(net_price,row)

cp(quantity,row)=quantity
pp(quantity_prop,row)=lag_quantity_prop
cp(net_price,row)=net_price

cp=column_as_is
pp=lag_column_as_is
row=nothing





```{r}
sales_wide_tbl |>
  # head(10) |>
  mutate(
    price=pr_delta*jan_quantity
    ,vol=quantity_delta*feb_price_realization
    ,jan_mix=jan_quantity/sum(jan_quantity)
    ,feb_mix=feb_quantity/sum(feb_quantity)
    ,mix_vol=(sum(feb_quantity)*jan_mix)-feb_quantity
    ,mix_price=jan_price_realization-mean(jan_price_realization)
    ,mix=mix_vol*mix_price
  ) |>
  summarise(
    vol=sum(vol,na.rm=TRUE)
    ,mix=sum(mix,na.rm=TRUE)
    ,price=sum(price,na.rm=TRUE)
    ,total_var=sum(net_price_delta,na.rm = TRUE)
  ) |>
  rowwise() |>
  mutate(
    total_fct=sum(c_across(c(-vol,-mix,-price)),na.rm = TRUE)
  )

```


```{r}




```{r}
factor(
  revenue~
      vol(vol     ~ pp(quantity,sum)          * d(price_realization) )+
    price(price   ~ d(quantity)           * cp(price_realization) )+
     mix1(mix1    ~ cp(quantity,sum))     * pp(mix)-cp(quantity))+
     mix2(mix2    ~ pp(price_realization) -  pp(price_realization,mean))+
      mix(mix     ~ mix1                  * mix2)
,time_unit="month"
)
```


```{r}

sales_wide_tbl <- sales_tbl |>
  group_split(month) |>
  map2(
    .y = c("jan", "feb"),
    .f = \(.x,.y) {
      x <- .x  # Assign .x to x for clarity
      y <- .y  # Assign .y to y for clarity

      x |>
        rename_with(
          .fn = ~ glue::glue("{y}_{.x}"),  # Use y and .col explicitly
          .cols = c(quantity, net_price, mean, price_realization,mix)
        ) |>
        select(-month)
    }
  ) |>
  reduce(
    \(.x,prev){
      .x |>
        left_join(
          prev
          ,by=join_by(product_key)
        )
    }
  ) |>
  mutate(
    quantity_delta=jan_quantity-feb_quantity
    ,net_price_delta=jan_net_price-feb_net_price
    ,mean_delta=jan_mean-feb_mean
    ,pr_delta=jan_price_realization-feb_price_realization
  )


## Methodology 1
## https://www.fpandaclub.com/handson/becoming-the-master-of-factor-analysis-of-profit-the-secrets-of-calculation

sales_wide_tbl |>
  # head(10) |>
  mutate(
    total_vol=sum(feb_quantity,na.rm = TRUE)
    ,jan_mix=jan_quantity/sum(jan_quantity)
    ,feb_mix=feb_quantity/sum(feb_quantity)
    ,vol_reclac=(total_vol*jan_mix)
    ,vol_effect=(vol_reclac*jan_price_realization)
    ,vol=vol_effect-jan_net_price
    ,mix_effect=(feb_quantity*jan_price_realization)
    ,mix=mix_effect-vol_effect
    ,price_effect=(feb_price_realization*feb_quantity)
    ,price=price_effect-mix_effect
  ) |>
  relocate(vol,mix,price,pr_delta,total_vol,feb_mix,vol_reclac) |>
  summarise(
    vol=sum(vol)
    ,mix=sum(mix,na.rm=TRUE)
    ,price=sum(price,na.rm=TRUE)
    ,total_var=sum(net_price_delta,na.rm = TRUE)
  ) |>
  rowwise() |>
  mutate(
    total_fct=sum(c_across(c(-vol,-mix,-price)),na.rm = TRUE)
  )
```


```{r}
library(tidyverse)


out <- formula(
  revenue~
      vol     ~ pp(quantity)          * d(price_realization) +
    price   ~ d(quantity)           * cp(price_realization)
)


formula(
  revenue~
      vol     ~ pp(quantity)          * d(price_realization) +
    price(price   ~ d(quantity)           * cp(price_realization) )+
     mix1(mix1    ~ cp(quantity,sum))     * pp(mix)-cp(quantity))+
     mix2(mix2    ~ pp(price_realization) -  pp(price_realization,mean))+
      mix(mix     ~ mix1                  * mix2)
# ,time_unit="month"
)

fpaR::sales |>
  fpaR::mom(date_var = order_date,value_var = quantity) |> tail()

str(formula_vec)
attr(formula_vec,"term.label")[[1]]
attr(formula_vec,"variables")
attr(formula_vec,"factors")

formula_vec[[2]][3]

price=pr_delta*jan_quantity
,vol=quantity_delta*feb_price_realization
,mix_vol=(sum(feb_quantity)*jan_mix)-feb_quantity
,mix_price=jan_price_realization-mean(jan_price_realization)
,mix=mix_vol*mix_price


rlang::fn_body(out)
```


```{r}

library(tidyverse)



```

I want a function in R that uses the below syntax
formula(net_price~cp(sales,quantity,'row') * pp(sales,quantity_prop,'row') - pp(sales,quantity,'row') *pp(sales,quantity,'row'))

where the final result will be a mutate function where net_price (to the left of ~) is the name of the column and each of 
 cp(sales,quantity,'row'), pp(sales,quantity_prop,'row'),pp(sales,quantity,'row') and pp(sales,quantity,'row') will be captured call objects and executed which will eventually return column names so that mutate will use the armethic indicators against the columns names. 
