---
title: "README"
format: gfm
---

# Developer Notes

## 2024-08-10

-   Cleaning up documentation
-   Make unit tests for yoy 
-   Clean up consistency of syntax
-   Fixed make aggregation tbl assertthat formula

## 2024-08-11

-   Cleaned up group by logic in time intelligence functions


## 2024-08-13
-   translated make_aggregation_tbl to sql (without group logic)
-   fixed how to add sub query that contains all previous steps
## 2024-08-15
-   fixed with and cte functions to build queries quicker

## 2024-08-31
-   I figured out template for factor analysis -- now i just need to figure out methodology and modeling approach
-   will use formula approach (target~factor1+factor2+factor3+...)
-   do pop calculations first with ti functions
-   then sequence the attribution
-   documented two methodologies for calculating sales price mix


## 2024-09-01
-   updated formula calls and structure of methodology -- period + agg args
-   established possible framework to execute 


## 2024-09-07

-   Finished with() and cte() helper functions for time intelligence conversions
-   pushed sql helper function to package
    
    -   sql_query_select
    -   date_seq_sql
    -   capture_original_query
    -   with()
    -   cte()
- started rewriting queries to take advantage of new sql helper functions
- create first draft of make_aggregation_tbl in sql -- need to work through groups version
    
    -   Error in referencing the column names (coming in quoted) 
    
## 2024-09-10

-   figured out quoting issue and it can be solved with DBI::Id("test", "species")

## 2024-09-11
-   Figure out how to do unique groups issue, so just need to optimize for multiple groups and clean up
-   Then can move to other time intelligence functions

2024-09-14

-  solved multiple group issue (finally!) need to clean up and layer it to R package and document / clean up 
- need add logic for one group var vs. multiple group var
    
## to do list

-   New Products introduced within the period
-   Discontinued Products within the period
-   Discounts and Returns for products that result in zero or negative revenue over a period
-   Figure out how to delay unquoted evaluation so that we can pass args to make aggregation tbl
-   replicate time intelligence functions
-   update examples with contoso example
-   covert lm to sql equivalents
-   design and create robust unit tests
-   make 554 calendar date table
-   create documentation with examples

# Introduction to fpaR

This is a collection of business intelligence tools and patterns to help with common data queries and insights. This is based on numerous blogs, books, post and work experience in what I have seen as most frequent analysis /work effort.


The package functions can be split between three focus areas:

-   Segmentation strategies

    -   UMAP
    -   Kmeans
    -   ABC
    -   New vs. Returning Customers/ Vendors
    
-  Time intelligence functions

    -   List of standard time intelligence functions such as month-to-date, year over year for standard and non-standard calendars
    -   See table before complete list
    -   Non-standard calendar such a 5-5-4 calendar

- DAX Equivalents

    -   Applied statsitical techniques to underestand variance analysis including graphing techniques
    -   Like for like comparison
        -   You calculate certain components such as price, customer turnover, high margin vs. jkdjklow margin segmentations
        
-   Factor / Variation analysis
  
    -   price, quantity, mix factor analysis with variations



The package is designed to work with regular tibble data and database by translating the functions to their sql variation


## Installation


You can install the development version of fpaR from [GitHub](https://github.com/alejandrohagan/fpaR) with:

```{r}
#| message: false
#| warning: false
#| error: false
#| eval: false


# install.packages("devtools")
devtools::install_github("alejandrohagan/fpaR")

```

## Package Components

This is correctly under development and is in early stages!

This package is heavily inspired by the DAX Patterns for the time intelligence functions

**Segmentation helpers:**

-   **make_segmentation()** will use kmeans and umap to create a segmentation
-   **make_cohort_tbl()** will segment your data into cohorts based on their first transaction date
-   **abc()** will create a three category segmentation ('A','B','C') based on the cut points that are submitted
-   **abc_graph()** will plot your `abc()` segmentation 

**Calculation Helpers**

-   **count_plus()** will replicate the handy dplyr::count() function but augments with portions and cumulative sum of proportions

-   **show_excel()** will show your tibbles or data frames steps in excel

-   **divide** will give an default  NA or alternative  to dividing by zero that isn't `Inf` (model after [DAX divide](https://learn.microsoft.com/en-us/dax/divide-function-dax))


**Time Intelligence Functions**

-   Standard time intelligence functions^[See table below]

    -   These are time intelligence fucntions that are consistent with a standard calendar (eg. the calendar on your computer or your phone)
-   Non-standard time intelligence functions
    
    -   These are calendars that maybe more common in the retail business (Eg. 5-4-5 quarter calcualtion)
    -   These calendars aim to control for weekends as that may be larger driver of sales

Time intelligence functions aren't always as straight forward as you want them to be as you may have missing dates from previous comparison periods.

The function basically ensures you have a complete date field and then does common time intelligence functions against the augmented data








**Variation Analysis**

-   variance / factor analysis

**Practice Datasets**

-   Microsoft's Contoso Dataset to help with practice with transaction data in both tibble and duckdb data


```{r}
#| echo: false
#| warning: false
#| message: false
#| error: false

tibble::tribble(
   ~"Short Name",~"Description",
    "YTD", 	"Year-to-date",	 
    "QTD", 	"Quarter-to-date",	  		 
    "MAT", 	"Moving annual total", 	  		 
    "PYC", 	"Previous year complete", 		  	 
    "PQC", 	"Previous quarter complete", 		  	 
    "PMC", 	"Previous month complete", 		  	 
    "PP", 	"Previous period; automatically selects year, quarter, or month" , 	 
    "PYMAT", 	"Previous year moving annual total" ,			 
    "MATG", 	"Moving annual total growth", 			
    "POP", 	"Period-over-period; automatically selects year, quarter, or month", 	  	  	
    "PYTD", 	"Previous year-to-date", 			 
    "PQTD",	"Previous quarter-to-date", 			 
    "YOYTD", 	"Year-over-year-to-date" ,			
    "QOQTD", "	Quarter-over-quarter-to-date", 			
    "YTDOPY", 	"Year-to-date-over-previous-year", 			
    "QTDOPQ", 	"Quarter-to-date-over-previous-quarter", 	
      
  )
```








