---
title: "README"
format: gfm
---
# fpaR


This is a collection of business intelligence tools and patterns to help with common data queries and insights. 
This is based on numerous blogs, books, post and work experience in what I have seen as most frequent analysis /work effort

The package functions can be split between three focus areas:

1) Segmentation strategies
    
    -   UMAP
    -   Kmeans
    -   ABC
2) Time intelligence functions

    -   List of standard time intelligence functions such as month-to-date, year over year for standard and non-standard calendars
    -   See table before complete list
3) Variance analysis strategies
    
    -   Applied statsitical techniques to underestand variance analysis including graphing techniques
        -   quantile regression
        -   anova 
    -   Like for like comparison

## Installation

You can install the development version of fpaR from [GitHub](https://github.com/) with:

```{r}
# install.packages("devtools")
devtools::install_github("alejandrohagan/fpaR")
```

This is correctly under development and is in early stages!

This package is heavily inspired by the DAX Patterns for the time intelligence functions


Segmentation helpers:
-   **make_segmentation()** will use kmeans and umap to create a segmentation
-   ***abc_segmentation()***
-   new vs. returning customers

Calculation Helpers
-   **count_plus()** will replicate the handy dplyr::count() function
-   **show_excel()** will show interim tranformation steps in excell
-   **divide** will give an alternative results to dividing by zero that isn't `Inf`


-   standard time intelligence functions
-   non-standard time intelligence functions
-   Transition metrics
-   variance / factor analysis
-   Practice datasets

```{r}

measurement_descriptions <- tibble::tribble(
  ~"Short Name",~"Description",
  "YTD", 	"Year-to-date",	 
"QTD", 	"Quarter-to-date",	  		 
"MAT", 	"Moving annual total", 	  		 
"PY", 	"Previous year", 		  	 
"PQ", 	"Previous quarter", 		  	 
"PM", 	"Previous month", 		  	 
"PYC", 	"Previous year complete", 		  	 
"PQC", 	"Previous quarter complete", 		  	 
"PMC", 	"Previous month complete", 		  	 
"PP", 	"Previous period; automatically selects year, quarter, or month" , 	 
"PYMAT", 	"Previous year moving annual total" ,			 
"YOY", 	"Year-over-year", 	  	  	
"QOQ", 	"Quarter-over-quarter", 	  	  	
"MOM", 	"Month-over-month",	  	  	
"MATG", 	"Moving annual total growth", 			
"POP", 	"Period-over-period; automatically selects year, quarter, or month", 	  	  	
"PYTD", 	"Previous year-to-date", 			 
"PQTD",	"Previous quarter-to-date", 			 
"YOYTD", 	"Year-over-year-to-date" ,			
"QOQTD", "	Quarter-over-quarter-to-date", 			
"YTDOPY", 	"Year-to-date-over-previous-year", 			
"QTDOPQ", 	"Quarter-to-date-over-previous-quarter", 	
  
)

measurement_descriptions %>% 
  gt::gt() %>% 
  gt::tab_style(
    location=gt::cells_column_labels(),
    style=gt::cell_fill("blue",alpha=.3)
    )

```

		



```{r}
x=100/0
divide <- function(numerator,denominator,alternative_result){
 
  result <- return(numerator/denominator) 
  
  if(is.infinite(x)){
  NA_integer_
}else{
  print(x)
}
}



df <- data.frame(ID = 1:5,
                 Name = c("John", "Jane", "Michael", "Samantha", "Tom"),
                 Age = c(25, 30, 35, 22, 27),
                 stringsAsFactors = FALSE)

# Split the dataframe by columns
df_list <- split(df, f = seq_along(df))





```


