---
title: "README"
format: gfm
---

# Developer Notes

## 2024-08-10

-   Cleaning up documentation
-   Make unit tests for yoy 
-   Clean up consistency of syntax
-   Fixed make aggregation tbl assertthat formula


# Introduction to fpaR

This is a collection of business intelligence tools and patterns to help with common data queries and insights. This is based on numerous blogs, books, post and work experience in what I have seen as most frequent analysis /work effort

The package functions can be split between three focus areas:

1)  Segmentation strategies

    -   UMAP
    -   Kmeans
    -   ABC
    -   New vs. Returning Customers/ Vendors
    
2)  Time intelligence functions

    -   List of standard time intelligence functions such as month-to-date, year over year for standard and non-standard calendars
    -   See table before complete list
    -   Non-standard calendar such a 5-5-4 calendar

3)  Variance analysis strategies

    -   Applied statsitical techniques to underestand variance analysis including graphing techniques
        -   quantile regression
        -   anova
    -   Like for like comparison
    - Something like a mix between marginalmeans, anova, lm and rq 
        -   You calculate certain components such as price, customer turnover, high margin vs. low margin segmentations
        -   Then do a quantile regression on the parts to see how there is difference between high margin and low margin indicators
        -   ideally we then have our factors and we can simply take the coefficients and see directly the components that is driving changes
            -   so we start at customer level, store level and product level attributes and other componets that when muliplied and added get what you are calcualted
            -price X vol =rev
            -avg.price X product mix
            -vol =product mix X quantity

## Installation

You can install the development version of fpaR from [GitHub](https://github.com/alejandrohagan/fpaR) with:

```{r}
#| message: false
#| warning: false
#| error: false
#| eval: false


# install.packages("devtools")
devtools::install_github("alejandrohagan/fpaR")

```

## Package Components

This is correctly under development and is in early stages!

This package is heavily inspired by the DAX Patterns for the time intelligence functions

**Segmentation helpers:**

-   **make_segmentation()** will use kmeans and umap to create a segmentation
-   **abc_segmentation()**
-   Turnover customers

**Calculation Helpers**

-   **count_plus()** will replicate the handy dplyr::count() function but augments with portions and cumulative sum of proportions

-   **show_excel()** will show your tibbles or data frames steps in excel

-   **divide** will give an default  NA or alternative  to dividing by zero that isn't `Inf` (model after [DAX divide](https://learn.microsoft.com/en-us/dax/divide-function-dax))

-   Calculate() model after DAX calculate, basically a supercharged sumif that allows you to filter data with or with any filter context

**Time Intelligence Functions**

-   Standard time intelligence functions^[See table below]

    -   These are time intelligence fucntions that are consistent with a standard calendar (eg. the calendar on your computer or your phone)
-   Non-standard time intelligence functions
    
    -   These are calendars that maybe more common in the retail business (Eg. 5-4-5 quarter calcualtion)
    -   These calendars aim to control for weekends as that may be larger driver of sales

**Variation Analysis**

-   variance / factor analysis

**Practice Datasets**

-   Microsoft's Contoso Dataset to help with practice with transaction data


```{r}
#| echo: false
#| warning: false
#| message: false
#| error: false

tibble::tribble(
   ~"Short Name",~"Description",
    "YTD", 	"Year-to-date",	 
    "QTD", 	"Quarter-to-date",	  		 
    "MAT", 	"Moving annual total", 	  		 
    "PYC", 	"Previous year complete", 		  	 
    "PQC", 	"Previous quarter complete", 		  	 
    "PMC", 	"Previous month complete", 		  	 
    "PP", 	"Previous period; automatically selects year, quarter, or month" , 	 
    "PYMAT", 	"Previous year moving annual total" ,			 
    "MATG", 	"Moving annual total growth", 			
    "POP", 	"Period-over-period; automatically selects year, quarter, or month", 	  	  	
    "PYTD", 	"Previous year-to-date", 			 
    "PQTD",	"Previous quarter-to-date", 			 
    "YOYTD", 	"Year-over-year-to-date" ,			
    "QOQTD", "	Quarter-over-quarter-to-date", 			
    "YTDOPY", 	"Year-to-date-over-previous-year", 			
    "QTDOPQ", 	"Quarter-to-date-over-previous-quarter", 	
      
  )
```

Like-for-like comparison

factor analysis






