---
title: "Execute time-intelligence or segments class objects to return the underlying transformed table"
pagetitle: "Execute time-intelligence or segments class objects to return the underlying transformed table — calculate"
---

::: {.qrtdown-source .dont-index}
Source: [`R/methods.R`](https://codeberg.org/usrbinr/ti/src/branch/main/R/methods.R)
:::

<p>The <code>calculate()</code> function takes an object created by a time function (like <code>ytd()</code>, <code>mtd()</code>, or <code>qtd()</code>) or a segment function (like <code>cohort()</code> or <code>abc()</code>) and executes the underlying transformation logic.
It translates the function blueprint into an actionable query, returning the final data table.
</p>

## Usage

```r
calculate(x, ...)

## S7 method for class <ti::ti>
calculate(x, ...)

## S7 method for class <ti::segment_cohort>
calculate(x, ...)

## S7 method for class <ti::segment_abc>
calculate(x, ...)
```

## Arguments

<dl class="qrtdown-arguments">
<dt id="arg-x"><code>x</code><a class="anchor" aria-label="anchor" href="#arg-x"></a></dt>
<dd><p>ti object</p></dd>
<dt id="arg-"><code>...</code><a class="anchor" aria-label="anchor" href="#arg-"></a></dt>
<dd><p>Not used.</p></dd>
</dl>

## Value

<p>dbi object
</p>

## Details

<p>The TI and segment functions in <strong>ti</strong>—such as <code>ytd()</code> or <code>cohort()</code> and others—are designed to be <strong>lazy and database-friendly</strong>.
They do not perform the heavy data transformation immediately.
Instead, they return a blueprint object (of class <code>ti</code>,<code>segment_abc</code> or <code>segment_cohort</code>) that contains all the parameters and logic needed for the calculation.
</p>
<p><strong><code>calculate()</code></strong> serves as the <strong>execution engine</strong>.
</p>
<p>When called, it interprets the blueprint and generates optimized R code or SQL code (using the <code>dbplyr</code> package) that is then executed efficiently on the data source, whether it's an in-memory <code>tibble</code> or a remote database backend (like <code>duckdb</code> or <code>snowflake</code>).
This approach minimizes data transfer and improves performance for large datasets.
</p>
<p>The resulting table will be sorted by the relevant date column to ensure the correct temporal ordering of the calculated metrics.
</p>

## Examples

```{r}
#| include: false
options(width = 80)
set.seed(1014)
.qrtdown_root <- local({
  p <- normalizePath(".", winslash = "/", mustWork = FALSE)
  while (!file.exists(file.path(p, "DESCRIPTION")) && dirname(p) != p) {
    p <- dirname(p)
  }
  p
})
if (requireNamespace("pkgload", quietly = TRUE) &&
    file.exists(file.path(.qrtdown_root, "DESCRIPTION"))) {
  pkgload::load_all(.qrtdown_root, export_all = FALSE, helpers = FALSE,
                    quiet = TRUE)
} else {
  library(ti)
}
```

```{r}
#| collapse: true
#| comment: "#>"
#| error: true

x <- ytd(contoso::sales,.date=order_date,.value=quantity,calendar_type="standard")
calculate(x)
```

