---
title: "Current period month-to-date"
pagetitle: "Current period month-to-date — mtd"
---

::: {.qrtdown-source .dont-index}
Source: [`R/time_intelligence_classes.R`](https://codeberg.org/usrbinr/ti/src/branch/main/R/time_intelligence_classes.R)
:::

<p>This calculates the monthly cumulative sum of targeted value using a standard or 5-5-4 calendar respecting
any groups that are passed through with <code><a href="https://dplyr.tidyverse.org/reference/group_by.html">dplyr::group_by()</a></code>
</p>
<p>Use <a href="calculate.qmd">calculate</a> to return the results
</p>

## Usage

```r
mtd(.data, .date, .value, calendar_type = "standard", fiscal_year_start = 1)
```

## Arguments

<dl class="qrtdown-arguments">
<dt id="arg-data"><code>.data</code><a class="anchor" aria-label="anchor" href="#arg-data"></a></dt>
<dd><p>tibble or dbi object (either grouped or ungrouped)</p></dd>
<dt id="arg-date"><code>.date</code><a class="anchor" aria-label="anchor" href="#arg-date"></a></dt>
<dd><p>the date column to group by. Must be a Date or POSIXt type.</p></dd>
<dt id="arg-value"><code>.value</code><a class="anchor" aria-label="anchor" href="#arg-value"></a></dt>
<dd><p>the value column to summarize. Must be numeric.</p></dd>
<dt id="arg-calendar-type"><code>calendar_type</code><a class="anchor" aria-label="anchor" href="#arg-calendar-type"></a></dt>
<dd><p>select either 'standard', '445', '454', or '544' calendar, see 'Details' for additional information</p></dd>
<dt id="arg-fiscal-year-start"><code>fiscal_year_start</code><a class="anchor" aria-label="anchor" href="#arg-fiscal-year-start"></a></dt>
<dd><p>integer 1-12, the month the fiscal year starts nearest to (default 1 = January). Only used with retail calendars ('445', '454', '544').</p></dd>
</dl>

## Value

<p>ti object
</p>

## Details

<ul>
<li>
<p> This function creates a complete calendar object that fills in any missing days, weeks, months, quarters, or years
</p>
</li>
<li>
<p> If you provide a grouped object with <code><a href="https://dplyr.tidyverse.org/reference/group_by.html">dplyr::group_by()</a></code>, it will generate a complete calendar for each group
</p>
</li>
<li>
<p> The function creates a <code>ti</code> object, which pre-processes the data and arguments for further downstream functions
</p>
</li>
</ul>

<h4><strong>NA Handling</strong></h4>
<ul>
<li>
<p> NA values in the <code>.value</code> column propagate through cumulative sums (cumsum returns NA once an NA is encountered)
</p>
</li>
<li>
<p> NA values in the <code>.date</code> column are excluded from the calendar join and will not appear in results
</p>
</li>
<li> 
</li>
<li>
<p> Missing dates (gaps in your data) are filled with 0 values, not NA
</p>
</li>
</ul>

<p><strong>standard calendar</strong>
</p>
<ul>
<li>
<p> The standard calendar splits the year into 12 months (with 28–31 days each) and uses a 7-day week
</p>
</li>
<li>
<p> It automatically accounts for leap years every four years to match the Gregorian calendar
</p>
</li>
</ul>

<p><strong>5-5-4 calendar</strong>
</p>
<ul>
<li>
<p> The 5-5-4 calendar divides the fiscal year into 52 weeks (occasionally 53), organizing each quarter into two 5-week periods and one 4-week period.
</p>
</li>
<li>
<p> This system is commonly used in retail and financial reporting
</p>
</li>
</ul>

## See Also

<p>Other time_intelligence:
<code><a href="atd.qmd">atd()</a></code>,
<code><a href="dod.qmd">dod()</a></code>,
<code><a href="mom.qmd">mom()</a></code>,
<code><a href="momtd.qmd">momtd()</a></code>,
<code><a href="mtdopm.qmd">mtdopm()</a></code>,
<code><a href="pmtd.qmd">pmtd()</a></code>,
<code><a href="pqtd.qmd">pqtd()</a></code>,
<code><a href="pwtd.qmd">pwtd()</a></code>,
<code><a href="pytd.qmd">pytd()</a></code>,
<code><a href="qoq.qmd">qoq()</a></code>,
<code><a href="qoqtd.qmd">qoqtd()</a></code>,
<code><a href="qtd.qmd">qtd()</a></code>,
<code><a href="qtdopq.qmd">qtdopq()</a></code>,
<code><a href="wow.qmd">wow()</a></code>,
<code><a href="wowtd.qmd">wowtd()</a></code>,
<code><a href="wtd.qmd">wtd()</a></code>,
<code><a href="wtdopw.qmd">wtdopw()</a></code>,
<code><a href="yoy.qmd">yoy()</a></code>,
<code><a href="yoytd.qmd">yoytd()</a></code>,
<code><a href="ytd.qmd">ytd()</a></code>,
<code><a href="ytdopy.qmd">ytdopy()</a></code>
</p>

## Examples

```{r}
#| include: false
options(width = 80)
set.seed(1014)
.qrtdown_root <- local({
  p <- normalizePath(".", winslash = "/", mustWork = FALSE)
  while (!file.exists(file.path(p, "DESCRIPTION")) && dirname(p) != p) {
    p <- dirname(p)
  }
  p
})
if (requireNamespace("pkgload", quietly = TRUE) &&
    file.exists(file.path(.qrtdown_root, "DESCRIPTION"))) {
  pkgload::load_all(.qrtdown_root, export_all = FALSE, helpers = FALSE,
                    quiet = TRUE)
} else {
  library(ti)
}
```

```{r}
#| collapse: true
#| comment: "#>"
#| error: true

mtd(contoso::sales,.date=order_date,.value=quantity,calendar_type="standard")
```

