---
title: "Current full period quarter over previous full period quarter"
pagetitle: "Current full period quarter over previous full period quarter — qoq"
---

::: {.qrtdown-source .dont-index}
Source: [`R/time_intelligence_classes.R`](https://codeberg.org/usrbinr/ti/src/branch/main/R/time_intelligence_classes.R)
:::

<ul>
<li>
<p> This calculates the full quarter value compared to the previous quarter value respecting
any groups that are passed through with <code><a href="https://dplyr.tidyverse.org/reference/group_by.html">dplyr::group_by()</a></code>
</p>
</li>
<li>
<p> Use <a href="calculate.qmd">calculate</a> to return the results
</p>
</li>
</ul>

## Usage

```r
qoq(
  .data,
  .date,
  .value,
  calendar_type = "standard",
  lag_n = 1,
  fiscal_year_start = 1
)
```

## Arguments

<dl class="qrtdown-arguments">
<dt id="arg-data"><code>.data</code><a class="anchor" aria-label="anchor" href="#arg-data"></a></dt>
<dd><p>tibble or dbi object (either grouped or ungrouped)</p></dd>
<dt id="arg-date"><code>.date</code><a class="anchor" aria-label="anchor" href="#arg-date"></a></dt>
<dd><p>the date column to group by. Must be a Date or POSIXt type.</p></dd>
<dt id="arg-value"><code>.value</code><a class="anchor" aria-label="anchor" href="#arg-value"></a></dt>
<dd><p>the value column to summarize. Must be numeric.</p></dd>
<dt id="arg-calendar-type"><code>calendar_type</code><a class="anchor" aria-label="anchor" href="#arg-calendar-type"></a></dt>
<dd><p>select either 'standard', '445', '454', or '544' calendar, see 'Details' for additional information</p></dd>
<dt id="arg-lag-n"><code>lag_n</code><a class="anchor" aria-label="anchor" href="#arg-lag-n"></a></dt>
<dd><p>the number of periods to lag</p></dd>
<dt id="arg-fiscal-year-start"><code>fiscal_year_start</code><a class="anchor" aria-label="anchor" href="#arg-fiscal-year-start"></a></dt>
<dd><p>integer 1-12, the month the fiscal year starts nearest to (default 1 = January). Only used with retail calendars ('445', '454', '544').</p></dd>
</dl>

## Value

<p>ti object
</p>

## See Also

<p>Other time_intelligence:
<code><a href="atd.qmd">atd()</a></code>,
<code><a href="dod.qmd">dod()</a></code>,
<code><a href="mom.qmd">mom()</a></code>,
<code><a href="momtd.qmd">momtd()</a></code>,
<code><a href="mtd.qmd">mtd()</a></code>,
<code><a href="mtdopm.qmd">mtdopm()</a></code>,
<code><a href="pmtd.qmd">pmtd()</a></code>,
<code><a href="pqtd.qmd">pqtd()</a></code>,
<code><a href="pwtd.qmd">pwtd()</a></code>,
<code><a href="pytd.qmd">pytd()</a></code>,
<code><a href="qoqtd.qmd">qoqtd()</a></code>,
<code><a href="qtd.qmd">qtd()</a></code>,
<code><a href="qtdopq.qmd">qtdopq()</a></code>,
<code><a href="wow.qmd">wow()</a></code>,
<code><a href="wowtd.qmd">wowtd()</a></code>,
<code><a href="wtd.qmd">wtd()</a></code>,
<code><a href="wtdopw.qmd">wtdopw()</a></code>,
<code><a href="yoy.qmd">yoy()</a></code>,
<code><a href="yoytd.qmd">yoytd()</a></code>,
<code><a href="ytd.qmd">ytd()</a></code>,
<code><a href="ytdopy.qmd">ytdopy()</a></code>
</p>

## Examples

```{r}
#| include: false
options(width = 80)
set.seed(1014)
.qrtdown_root <- local({
  p <- normalizePath(".", winslash = "/", mustWork = FALSE)
  while (!file.exists(file.path(p, "DESCRIPTION")) && dirname(p) != p) {
    p <- dirname(p)
  }
  p
})
if (requireNamespace("pkgload", quietly = TRUE) &&
    file.exists(file.path(.qrtdown_root, "DESCRIPTION"))) {
  pkgload::load_all(.qrtdown_root, export_all = FALSE, helpers = FALSE,
                    quiet = TRUE)
} else {
  library(ti)
}
```

```{r}
#| collapse: true
#| comment: "#>"
#| error: true

qoq(contoso::sales,.date=order_date,.value=quantity,calendar_type='standard',lag_n=1)
```

