---
title: "Working with Groups"
subtitle: "How grouping changes every calculation in the package"
author: "Alejandro Hagan"
format: html
---

```{r}
#| label: setup
#| include: false

library(ti)
library(dplyr)
```

## Grouping is the main dial

Every time-intelligence and segmentation function in `ti` respects `dplyr::group_by()`. There is no `.by` argument and no `group` parameter — you group the table before you call the function, and the calculation follows.

Ungrouped, `mtd()` gives you one month-to-date series for the whole company:

```{r}
#| label: ungrouped

contoso::sales |>
  mtd(.date = order_date, .value = gross_margin, calendar_type = "standard") |>
  calculate() |>
  collect() |>
  arrange(date) |>
  select(date, gross_margin, mtd_gross_margin) |>
  head(5)
```

Grouped by store, you get one independent series **per store**:

```{r}
#| label: grouped

by_store <- contoso::sales |>
  group_by(store_key) |>
  mtd(.date = order_date, .value = gross_margin, calendar_type = "standard") |>
  calculate() |>
  collect()

by_store |>
  arrange(store_key, date) |>
  select(store_key, date, gross_margin, mtd_gross_margin) |>
  head(5)
```

The grouping columns are carried through into the result, so they are available to filter and join on afterwards.

## The cumulative sum resets per group

This is the behaviour that makes grouping worth doing. Each group's running total is independent — store 10's month-to-date figure never picks up store 20's revenue.

```{r}
#| label: independent-series

by_store |>
  filter(year == 2023, month == 6) |>
  arrange(store_key, date) |>
  summarise(
    month_total = max(mtd_gross_margin),
    .by = store_key
  ) |>
  arrange(desc(month_total)) |>
  head(5)
```

## Multiple grouping levels

Group by as many columns as you need. The calculation is performed for every distinct combination:

```{r}
#| label: multi-level

contoso::sales |>
  group_by(store_key, product_key) |>
  mtd(.date = order_date, .value = gross_margin, calendar_type = "standard") |>
  calculate() |>
  collect() |>
  select(store_key, product_key, date, mtd_gross_margin) |>
  head(5)
```

Be aware of the cost. Because `ti` completes the calendar for each group (see [Calendar Completion and Missing Dates](missing_dates.qmd)), the row count is roughly *number of groups × number of days in that group's active range*. Grouping by two high-cardinality columns can multiply your result set considerably — one reason to keep the pipeline lazy and filter in the database.

## Each group gets its own calendar range

A subtle and important point: `ti` does not force every group onto the same date range. Each group's calendar is completed only within **that group's own** first and last observed date.

Consider two stores, where store B opens two months after store A:

```{r}
#| label: per-group-bounds

openings <- tibble::tibble(
  store      = c("A", "A", "A", "B", "B"),
  order_date = as.Date(c("2024-01-01", "2024-01-03", "2024-03-01",
                         "2024-03-01", "2024-03-05")),
  amount     = c(10, 20, 30, 40, 50)
)

openings |>
  group_by(store) |>
  ytd(.date = order_date, .value = amount, calendar_type = "standard") |>
  calculate() |>
  collect() |>
  summarise(
    first_date = min(date),
    last_date  = max(date),
    n_rows     = n(),
    .by = store
  ) |>
  arrange(store)
```

Store B's series starts in March, not January. It is **not** back-filled with two months of zeroes just because store A had data then.

This is the right default for retail and subscription analysis. A store that opened in March did not have a bad January — it had no January. Padding it with zeroes would drag down every average and make year-over-year comparisons meaningless.

## Checking your grouping before you run

Because a `ti` object is a blueprint, you can inspect what it is about to do. Printing it reports the group count:

```{r}
#| label: print-groups

contoso::sales |>
  group_by(store_key) |>
  mtd(.date = order_date, .value = gross_margin, calendar_type = "standard")
```

If the group count is not what you expected, you have caught the mistake before running a query over the full table.

## Grouping and segmentation

The segmentation functions treat grouping differently: for `abc()`, the grouping columns define **what is being ranked**. Grouping by store ranks stores against each other by their contribution to the total.

```{r}
#| label: abc-grouping

contoso::sales |>
  group_by(store_key) |>
  abc(category_values = c(0.7, 0.96, 1), .value = gross_margin) |>
  calculate() |>
  collect() |>
  select(store_key, abc_gross_margin, prop_total, cum_prop_total, category_name) |>
  head(5)
```

Change the grouping column and you change the question: group by `product_key` and you are asking which products drive margin, not which stores.

## Key points

- Group with `dplyr::group_by()` before calling the function; there is no `.by` argument.
- Cumulative sums reset per group and never bleed across groups.
- Grouping columns are carried into the result.
- Each group's calendar is completed within its own date range, so late-starting groups are not zero-padded backwards.
- Print the object to check the group count before executing.
- For `abc()`, the grouping columns define the population being ranked.
