---
title: "A Month-End Close, End to End"
subtitle: "Composing several time-intelligence functions into one reporting pack"
author: "Alejandro Hagan"
format: html
---

```{r}
#| label: setup
#| include: false

library(ti)
library(dplyr)
library(gt)
```

## The task

Every other article in this package demonstrates one function at a time. Real reporting does not work that way. At month end you are asked for a single table that answers several questions at once:

- How did each store perform this month to date?
- How does that compare with the same point last month?
- Where does the year stand, and how does that compare with last year?

That is four metrics — `mtd`, `pmtd`, `ytd`, `pytd` — from three function calls, combined into one pack. This article builds it.

We will close on **30 June 2023**, at store level, measuring gross margin.

```{r}
#| label: parameters

close_date <- as.Date("2023-06-30")
```

## One metric at a time

Each call follows the same shape: group, calculate, collect. Keep only the identifying columns and the metric, so the pieces join cleanly.

Month to date, and the prior month at the same point:

```{r}
#| label: mtd-pieces

mtd_tbl <- contoso::sales |>
  group_by(store_key) |>
  mtd(.date = order_date, .value = gross_margin, calendar_type = "standard") |>
  calculate() |>
  collect() |>
  select(store_key, date, mtd = mtd_gross_margin)

pmtd_tbl <- contoso::sales |>
  group_by(store_key) |>
  pmtd(.date = order_date, .value = gross_margin, calendar_type = "standard", lag_n = 1) |>
  calculate() |>
  collect() |>
  select(store_key, date, pmtd = pmtd_gross_margin)
```

`yoytd()` is the efficient choice for the annual view, because it returns **both** the current year to date and the prior year to date in one pass:

```{r}
#| label: ytd-piece

ytd_tbl <- contoso::sales |>
  group_by(store_key) |>
  yoytd(.date = order_date, .value = gross_margin, calendar_type = "standard", lag_n = 1) |>
  calculate() |>
  collect() |>
  select(store_key, date, ytd = ytd_gross_margin, pytd = pytd_gross_margin)
```

::: callout-tip
## Which annual function?

`yoytd()` compares year to date against the *prior year to date* — the equivalent point last year. `ytdopy()` compares year to date against the *full prior year*. For a close pack you almost always want `yoytd()`; comparing six months of this year against twelve months of last year will always look like a collapse.
:::

## Joining the pieces

Join on **both** `store_key` and `date`. This is the part worth getting right.

```{r}
#| label: join

pack <- mtd_tbl |>
  inner_join(pmtd_tbl, by = c("store_key", "date")) |>
  inner_join(ytd_tbl,  by = c("store_key", "date"))

nrow(pack)
```

Joining on `store_key` alone would silently produce a cross join of every date against every date. Joining on `date` alone would mix stores together. The pair is the grain of every result `ti` returns for a grouped calculation, so it is the correct join key.

## Selecting the close date and deriving variances

With the metrics aligned, the close pack is one `filter()` and a few `mutate()` calls:

```{r}
#| label: close-pack

close_pack <- pack |>
  filter(as.Date(date) == close_date) |>
  mutate(
    mtd_variance     = mtd - pmtd,
    mtd_variance_pct = if_else(pmtd == 0, NA, (mtd - pmtd) / pmtd),
    ytd_variance     = ytd - pytd,
    ytd_variance_pct = if_else(pytd == 0, NA, (ytd - pytd) / pytd)
  ) |>
  arrange(desc(ytd))

close_pack |>
  select(store_key, mtd, pmtd, mtd_variance, ytd, pytd, ytd_variance_pct) |>
  head(5)
```

Guard the percentage variances against a zero denominator. A store with no margin in the comparison period is a real and common case, and dividing by it produces `Inf` rather than an error — which will propagate quietly into a formatted table.

## Formatting the pack

```{r}
#| label: tbl-close-pack
#| tbl-cap: "Store performance as at 30 June 2023, ranked by year to date gross margin"

close_pack |>
  head(10) |>
  select(store_key, mtd, pmtd, mtd_variance, ytd, pytd, ytd_variance_pct) |>
  gt() |>
  tab_header(
    title    = md("**Month-End Close — Gross Margin**"),
    subtitle = "Top 10 stores as at 30 June 2023"
  ) |>
  tab_spanner(label = "Month to date", columns = c(mtd, pmtd, mtd_variance)) |>
  tab_spanner(label = "Year to date",  columns = c(ytd, pytd, ytd_variance_pct)) |>
  cols_label(
    store_key        = "Store",
    mtd              = "Current",
    pmtd             = "Prior",
    mtd_variance     = "Variance",
    ytd              = "Current",
    pytd             = "Prior",
    ytd_variance_pct = "Variance %"
  ) |>
  fmt_number(columns = c(mtd, pmtd, mtd_variance, ytd, pytd), decimals = 0, use_seps = TRUE) |>
  fmt_percent(columns = ytd_variance_pct, decimals = 1) |>
  sub_missing(missing_text = "—") |>
  tab_style(
    style     = cell_text(color = "#b8291f"),
    locations = cells_body(columns = mtd_variance, rows = mtd_variance < 0)
  ) |>
  tab_style(
    style     = cell_text(color = "#b8291f"),
    locations = cells_body(columns = ytd_variance_pct, rows = ytd_variance_pct < 0)
  )
```

## Reading the result

The pack separates two different signals that are easy to conflate:

- **Month-to-date variance** is a short-horizon operational signal. It is noisy — one large order lands or does not.
- **Year-to-date variance** is the trend. A store can be down on the month and comfortably up on the year.

Sorting by year to date and reading the month column across gives you the stores worth asking about: strong on the year, weak on the month.

## Keeping it lazy for larger data

This article collects each metric before joining, which is fine for the Contoso sample. On a warehouse table you would keep everything lazy and let the database perform the joins:

```{r}
#| label: lazy-pattern
#| eval: false

mtd_lazy <- db_sales |>
  group_by(store_key) |>
  mtd(order_date, gross_margin, "standard") |>
  calculate()                                   # <1>

pmtd_lazy <- db_sales |>
  group_by(store_key) |>
  pmtd(order_date, gross_margin, "standard", lag_n = 1) |>
  calculate()

mtd_lazy |>
  inner_join(pmtd_lazy, by = c("store_key", "date")) |>   # <2>
  filter(date == close_date) |>
  collect()                                                # <3>
```

1. No `collect()` — this stays a lazy `tbl_dbi`.
2. The join is compiled to SQL and executed by the database.
3. One `collect()` at the very end, on the filtered result.

See [Working with Databases](databases.qmd) for more on the lazy pipeline and inspecting the generated SQL.

## Key points

- Group, calculate, collect — then join. The join key for a grouped result is the grouping columns **plus** `date`.
- `yoytd()` returns current and prior year to date together; prefer it over separate `ytd()` and `pytd()` calls.
- Use `yoytd()` rather than `ytdopy()` for close packs; the latter compares against the full prior year.
- Guard percentage variances against a zero comparison period.
- On large data, join lazily and `collect()` once at the end.
