---
title: "Working with Databases"
subtitle: "How ti builds SQL, and why you should collect() last"
author: "Alejandro Hagan"
format: html
---

```{r}
#| label: setup
#| include: false

library(ti)
library(dplyr)
library(DBI)
```

## Everything is lazy

`ti` never touches your data when you call a time-intelligence function. `ytd()`, `mtd()`, `yoy()` and the rest all return a **blueprint** — an object describing the calculation to run. Nothing executes until you ask for it.

There are three distinct stages:

| Stage | Call | What happens |
|---|---|---|
| **Describe** | `mtd(...)` | Returns a `ti` object. No query, no data movement. |
| **Compile** | `calculate()` | Builds the query. Still lazy — a `tbl_dbi`, not a tibble. |
| **Execute** | `dplyr::collect()` | Runs the query and pulls the result into R. |

This matters because the expensive part — the windowed aggregation — happens **in the database**, not in R.

## Connecting to a database

`ti` works on any `dbplyr`-backed table. Here is a DuckDB connection with the Contoso sales data loaded into it:

```{r}
#| label: connect

con <- dbConnect(duckdb::duckdb())
dbWriteTable(con, "sales", contoso::sales)

db_sales <- tbl(con, "sales")

db_sales
```

The syntax from here is identical to the tibble syntax used everywhere else in these articles:

```{r}
#| label: db-mtd

result <- db_sales |>
  mtd(.date = order_date, .value = gross_margin, calendar_type = "standard")

class(result)
```

Still no query. `calculate()` compiles it:

```{r}
#| label: db-calculate

compiled <- calculate(result)

class(compiled)
```

Note that `compiled` is a `tbl_dbi` — a lazy database table. The data is still in DuckDB.

## Seeing the generated SQL

Because the result is a lazy `dbplyr` table, `dplyr::show_query()` will show you exactly what `ti` wrote:

```{r}
#| label: show-query

compiled |> show_query()
```

The period-to-date calculation compiles down to a **SQL window function** partitioned by the period columns. `ti` is not looping in R and it is not pulling your table into memory — it is writing the query you would have written by hand.

## Collect last, not first

The single most important habit: **keep the pipeline lazy for as long as possible and call `collect()` at the very end.**

```{r}
#| label: collect-last

compiled |>
  filter(year == 2023, month == 6) |>   # <1>
  arrange(date) |>
  collect() |>                          # <2>
  head(5)
```

1. Filtering happens *in the database*, on the full result set.
2. Only the filtered rows cross the boundary into R.

If you had called `collect()` before the `filter()`, DuckDB would have handed R every row and then R would have thrown most of them away.

## Tibbles get a database too

You do not need to set up a connection to benefit from this. When you pass a plain `data.frame` or tibble, `ti` registers it into a temporary in-memory DuckDB instance with `make_db_tbl()`:

```{r}
#| label: auto-duckdb

contoso::sales |>
  mtd(order_date, gross_margin, "standard") |>
  calculate() |>
  class()
```

The result is a `tbl_duckdb_connection`, not a tibble — even though the input was a tibble. Registration uses DuckDB's `duckdb_register()`, which is a virtual registration rather than a physical copy, so it is fast and does not duplicate your data.

This is why every example in these articles ends with `dplyr::collect()`.

## Dialect support

Date arithmetic differs across SQL engines, so `ti` detects the backend from the connection and generates the appropriate expression. Three dialects are recognised explicitly:

| Backend | Status |
|---|---|
| DuckDB | Recognised |
| Snowflake | Recognised |
| Postgres | Recognised |
| Anything else | Falls back to the DuckDB dialect |

The fallback is often fine, since many engines accept DuckDB-compatible date arithmetic. But it is a fallback, not a guarantee — if you are running `ti` against a backend not in the list above, verify the output of `show_query()` before trusting the numbers.

```{r}
#| label: disconnect
#| include: false

dbDisconnect(con, shutdown = TRUE)
```

## Key points

- Time-intelligence functions describe work; `calculate()` compiles it; `collect()` runs it.
- Use `show_query()` to inspect the SQL — period-to-date becomes a window function.
- Filter, join and arrange *before* `collect()` so the database does the work.
- Tibbles are transparently promoted to DuckDB, so the return value is always a lazy table.
- DuckDB, Snowflake and Postgres get dialect-specific date arithmetic; other backends fall back to DuckDB's.
