---
title: "Calendar Completion and Missing Dates"
subtitle: "Why ti fills your calendar, and what it refuses to fill"
author: "Alejandro Hagan"
format: html
---

```{r}
#| label: setup
#| include: false

library(ti)
library(dplyr)
```

## The problem with gaps

Transactional data has holes in it. Stores close on holidays, no one orders on a Sunday, a product sells nothing for a week. Your table has no row for those days — not a row with a zero, no row at all.

That is fine until you compare periods. Consider a week of margin data with three days missing:

```{r}
#| label: gappy-data

gappy <- tibble::tibble(
  order_date = as.Date(c("2024-01-01", "2024-01-03", "2024-01-06",
                         "2024-01-07", "2024-01-08", "2024-01-09",
                         "2024-01-11")),
  amount     = c(1200, 1100, 1300, 900, 1200, 850, 1450)
)

gappy
```

If you reach for `dplyr::lag()` to compute a day-over-day change, you get the wrong answer:

```{r}
#| label: lag-is-wrong

gappy |>
  mutate(
    previous_row = lag(amount),
    change       = amount - previous_row
  )
```

The row labelled `2024-01-03` reports a change against `2024-01-01`. But those are not consecutive days — January 2nd is missing. `lag()` gives you the **previous row**, and you asked for the **previous day**. On a table with gaps those are different questions.

The problem compounds as the gaps widen. `2024-01-06` is compared against `2024-01-03`, three days earlier.

## What ti does instead

Every `ti` calculation begins by building a complete calendar for the requested time unit, then joining your data onto it. Missing days appear as rows with a zero value:

```{r}
#| label: ti-fills

filled <- gappy |>
  dod(.date = order_date, .value = amount, calendar_type = "standard") |>
  calculate() |>
  collect() |>
  arrange(date)

filled |>
  select(date, amount, dod_amount, missing_date_indicator)
```

Now January 2nd exists, holds a zero, and the day-over-day comparison lines up with the actual calendar.

## The `missing_date_indicator` column

Zero-filling is useful for arithmetic and dangerous for interpretation. A day with no sales and a day with genuinely zero margin look identical once both are zeroes.

So `ti` labels them. Every completed calendar carries a `missing_date_indicator` column: `1` if the row was invented to fill a gap, `0` if it came from your data.

```{r}
#| label: indicator

filled |>
  count(missing_date_indicator)
```

Use it to keep the two cases apart. To count only trading days:

```{r}
#| label: trading-days

filled |>
  filter(missing_date_indicator == 0) |>
  summarise(
    trading_days = n(),
    mean_amount  = mean(amount)
  )
```

Compare that against the naive average over all rows, which is dragged toward zero by the padding:

```{r}
#| label: naive-average

filled |>
  summarise(
    all_rows    = n(),
    mean_amount = mean(amount)
  )
```

Both numbers are legitimate; they answer different questions. Averaging over the calendar tells you daily run rate including closures. Averaging over trading days tells you how a trading day performs. The indicator lets you choose deliberately rather than by accident.

## What ti refuses to fill

Calendar completion is bounded. `ti` fills gaps *inside* the observed date range and does not extend beyond it. There is no back-filling to the start of the year and no padding forward to today.

```{r}
#| label: bounded

filled |>
  summarise(first_row = min(date), last_row = max(date))
```

The series runs from the first observed date to the last, and no further. Data you never had does not get invented.

This bounding is applied **per group**. Each group's calendar is completed within that group's own first and last observation, which matters whenever groups enter or leave at different times:

```{r}
#| label: per-group

openings <- tibble::tibble(
  store      = c("A", "A", "A", "B", "B"),
  order_date = as.Date(c("2024-01-01", "2024-01-03", "2024-03-01",
                         "2024-03-01", "2024-03-05")),
  amount     = c(10, 20, 30, 40, 50)
)

openings |>
  group_by(store) |>
  ytd(.date = order_date, .value = amount, calendar_type = "standard") |>
  calculate() |>
  collect() |>
  summarise(first_date = min(date), last_date = max(date), n_rows = n(), .by = store) |>
  arrange(store)
```

Store B opens in March and its series starts in March. It is not given two months of zeroes to match store A. A store that opened in March did not have a bad January — it had no January, and padding it would corrupt every average and year-over-year comparison that touches it.

## Inspecting gaps before you calculate

A `ti` object reports what it found before you run anything. Printing one summarises the calendar it intends to build, including how many dates were missing:

```{r}
#| label: print-object

gappy |>
  dod(.date = order_date, .value = amount, calendar_type = "standard")
```

If the missing-date count is far higher than you expect, that is usually a signal about the data rather than the calculation — a wrong grouping column, a filter applied too early, or a date column that is not the one you meant.

## A related tool: calendar attributes

`augment_standard_calendar()` solves a different problem and is worth knowing about. It does **not** fill gaps — it annotates the rows you already have with calendar context: period boundaries, day counts, and how far through each period a date sits.

```{r}
#| label: augment

augmented <- gappy |>
  augment_standard_calendar(.date = order_date) |>
  collect()

augmented |>
  select(order_date, month_end_date, days_complete_in_month,
         days_remaining_in_month, weekend_indicator) |>
  head(5)
```

These columns are what you need for pacing and run-rate work — projecting a partial month to a full month, or explaining a shortfall by weekend placement rather than performance. The full set covers year, quarter, month and week boundaries, day-of-week and month labels, and complete/remaining day counts for each period.

## Key points

- `dplyr::lag()` gives you the previous **row**; period comparisons need the previous **date**. On gappy data these differ.
- `ti` completes the calendar before calculating, so comparisons align to real dates.
- `missing_date_indicator` marks invented rows — filter on it to separate closures from genuine zeroes.
- Completion is bounded by the observed range and never extends past it.
- Bounds are per group, so late-starting groups are not zero-padded backwards.
- Print the object to see the missing-date count before running the query.
